---
title: "final_analysis"
output: html_document
date: '2023-03-11'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


```{r cars}
library(tidyverse)
library(MatchIt)
library(modelsummary)
library(plm)
```

```{r}
df <- read.csv("nhanes_mortality.csv")

df <- df %>% drop_na(Gender)
df <- replace(df, is.na(df), 0) #replace all NAs with zero
sum(is.na(df))  #0

view(df)
str(df)
head(df)
df <- df %>% mutate(Gender = ifelse(Gender =="female", 1, 0))
df <- df %>% mutate(Have_insurance = ifelse(Have_insurance =="Yes", 1, 0))

df$Marriage<-as.factor(df$Marriage)
df$Education <- as.factor(df$Education)
df$Familty_Poverty_income_ratio <- as.factor(df$Familty_Poverty_income_ratio)
df$Ethnicity <- as.factor(df$Ethnicity)

df$BMI_category<-as.factor(df$BMI_category)
df$alcohol<-as.factor(df$alcohol)
df$smoke <- as.factor(df$smoke)
df$physical_activity <- as.factor(df$physical_activity)
df$CVD <- as.factor(df$CVD)
df$smoke <- as.factor(df$smoke)
df$year <- as.factor(df$Year)
df$Cancer <- as.factor(df$Cancer)
df$mortstat <- as.factor(df$mortstat)
df$diabetes.y <- as.factor(df$diabetes.y)

sum(is.na(df))  #0


table(df$Ethnicity)

class(df$diabetes.y)

#view(df)

df2007 <- df %>% filter(year == 2007) #%>% filter(Age %in% (50:54))
df2009 <- df %>% filter(year == 2009) #%>% filter(Age %in% (52:56))
df2011 <- df %>% filter(year == 2011) #%>% filter(Age %in% (54:58))
df2013 <- df %>% filter(year == 2013) #%>% filter(Age %in% (56:60))
df2015 <- df %>% filter(year == 2015) #%>% filter(Age %in% (58:62))
df2017 <- df %>% filter(year == 2017) #%>% filter(Age %in% (60:64))

#match within each year

#analysis_match <- matchit(Have_insurance ~ Gender + Familty_Poverty_income_ratio + Age + Ethnicity +
#                            Marriage + Education + CVD + Cancer + alcohol + smoke + physical_activity + BMI_category,
#                            data = df,
#                            method = "nearest", 
#                          weights = weights,
#                           distance='scaled_euclidean', # metric for distance
#                     replace=TRUE, # do we use replacements in matches?
#                     verbose=TRUE)

analysis_match_2007 <- matchit(Have_insurance ~ Gender + Familty_Poverty_income_ratio + Age + Ethnicity +
                            Marriage + Education + CVD + Cancer + alcohol + smoke + physical_activity + BMI_category,
                            data = df2007,
                             method = "exact", 
                     verbose=TRUE)

analysis_match_2009 <- matchit(Have_insurance ~ Gender + Familty_Poverty_income_ratio + Age + Ethnicity +
                            Marriage + Education + CVD + Cancer + alcohol + smoke + physical_activity + BMI_category,
                            data = df2009,
                             method = "exact", 
                     verbose=TRUE)


analysis_match_2011 <- matchit(Have_insurance ~ Gender + Familty_Poverty_income_ratio + Age + Ethnicity +
                            Marriage + Education + CVD + Cancer + alcohol + smoke + physical_activity + BMI_category,
                            data = df2011,
                             method = "exact", 
                     verbose=TRUE)


analysis_match_2013 <- matchit(Have_insurance ~ Gender + Familty_Poverty_income_ratio + Age + Ethnicity +
                            Marriage + Education + CVD + Cancer + alcohol + smoke + physical_activity + BMI_category,
                            data = df2013,
                             method = "exact", 
                     verbose=TRUE)


analysis_match_2015 <- matchit(Have_insurance ~ Gender + Familty_Poverty_income_ratio + Age + Ethnicity +
                            Marriage + Education + CVD + Cancer + alcohol + smoke + physical_activity + BMI_category,
                            data = df2015,
                             method = "exact", 
                     verbose=TRUE)


analysis_match_2017 <- matchit(Have_insurance ~ Gender + Familty_Poverty_income_ratio + Age + Ethnicity +
                            Marriage + Education + CVD + Cancer + alcohol + smoke + physical_activity + BMI_category,
                            data = df2017,
                            method = "exact", 
                     verbose=TRUE)


#summary(analysis_match)
#analysis_match_data <-  match.data(analysis_match)

summary(analysis_match_2007)
summary(analysis_match_2009)
summary(analysis_match_2011)
summary(analysis_match_2013)
summary(analysis_match_2015)
summary(analysis_match_2017)


analysis_match_2007_data <-  match.data(analysis_match_2007)
analysis_match_2009_data <-  match.data(analysis_match_2009)
analysis_match_2011_data <-  match.data(analysis_match_2011)
analysis_match_2013_data <-  match.data(analysis_match_2013)
analysis_match_2015_data <-  match.data(analysis_match_2015)
analysis_match_2017_data <-  match.data(analysis_match_2017)

analysis_match_data_merged<-rbind(analysis_match_2007_data,analysis_match_2009_data,analysis_match_2011_data,analysis_match_2013_data,analysis_match_2015_data,analysis_match_2017_data)

view(analysis_match_data_merged)

```

```{r message=FALSE, warning=FALSE}


#all_death_regression <-glm(mortstat ~ Have_insurance + Have_insurance:Age + year + Age, data=analysis_match_data_merged, family='binomial'(link='logit'), weights=weights)
#family=quasibinomial
#all_death_regression <-glm(mortstat ~ diabetes.x+Have_insurance + Have_insurance:Age + year + Age, data=analysis_match_data_merged, family='quasibinomial'(link='logit'), weights=weights)
#summary(all_death_regression)

class(df$year)

#diabetes_death_regression <-glm(diabetes.y ~ Have_insurance + Have_insurance:Age + year + Age, data=analysis_match_data_merged, family='binomial'(link='logit'), weights=weights)
#summary(diabetes_death_regression)
#family=quasibinomial
#diabetes_death_regression <-glm(diabetes.y ~ Have_insurance + Have_insurance:Age + year + Age, data=analysis_match_data_merged, family='quasibinomial'(link='logit'), weights=weights)


#first difference models
all_death_regression_fd <-  plm(mortstat ~ diabetes.x+Have_insurance + Have_insurance:Age + year:Have_insurance + year + Age, data = analysis_match_data_merged,index = "year", model = "fd")
summary(all_death_regression_fd)


diabetes_death_regression_fd <-  plm(diabetes.y ~ Have_insurance + Have_insurance:Age + year + Age, data=analysis_match_data_merged,index = c("SEQN", "year"), model = "fd")
summary(diabetes_death_regression_fd)


#demean models
all_death_regression_fe <-  plm(mortstat ~ diabetes.x+Have_insurance + Have_insurance:Age + year:Have_insurance + year + Age, data = analysis_match_data_merged,index = c("SEQN", "year"), model = "within")
summary(all_death_regression_fe)

diabetes_death_regression_fe <-  plm(diabetes.y ~ Have_insurance + Have_insurance:Age + year + Age, data=analysis_match_data_merged,index = c("SEQN", "year"), model = "within")
summary(diabetes_death_regression_fe)







msummary(list("all_death_fd"=all_death_regression_fd, "diabetes_death_fd" = diabetes_death_regression_fd,"all_death_fe"=all_death_regression_fe, "diabetes_death_fe" = diabetes_death_regression_fe),
         stars=c('*' = .1, '**' = .05, '***' = .01))

```

### use fixef
### should we use fixed effect or first difference
### what regression model - glm or plm?
### exact matching or nearest matching - ask petros
### matching by year - should we just seperate the waves within the dataset ; done and confirm wiht petros
### quasibinomial vs binomial for the glm regression; ask petros/look up
### private vs no insurance, public vs no insurance - Guan
### match private vs no, match public vs no
### diagnostic test and assumption test
### look at interval censoring

```{r}
#Interval sensoring






```







